Experiences with Planning for Natural Language Generation

### 2009-05-07 trunk/Papers/ci-spark/domains.tex
started revising the crisp presentation

Although the sentence generation task is often split up into two
separate steps for sentence planning and surface realization, the
decisions that an NLG system must make in these two modules often
interact. For instance, in order to generate \emph{referring
  expressions (REs)}, i.e.\ unique descriptions of individuals in the
universe of discourse, it is necessary to know what properties of
individuals can be expressed as nouns and adjectives. Grammatical
information like this should belong exclusively to the surface
realization module; but on the other hand, RE generation enriches the
input meaning representation and is therefore usually seen as part of
sentence planning.

One seminal approach to solving the sentence planning and surface
realization problems together is offered by the SPUD generation system
(``Sentence Planning Using Description''; \citealt{Stone2003a}). This
approach assumes a tree-adjoining grammar (TAG) \citep{joshi;etal1997}
whose lexical entries are equipped with semantic and pragmatic
information, as well as two sets of ground atoms: the
\emph{communicative goal} that specifies what semantic information we
must express, and a \emph{knowledge base}. Let's consider an example
to illustrate the problem that SPUD solves. Consider a knowledge base
containing the individuals $e$, $r_1$ and $r_2$, and a set of
attributes encoding the fact that $r_1$ and $r_2$ are rabbits, $r_1$
is white and $r_2$ is brown, and $e$ is an event in which $r_1$
sleeps.  Say that we want to express the information
$\{\mathsf{sleep}(e,r_1)\}$ using the tree-adjoining grammar shown in
Figure~\ref{fig:white-rabbit-sleeps-grammar}. This grammar consists of
\emph{elementary trees} (i.e., the disjoint trees in the figure), each
of which contributes certain \emph{semantic content}. We can
instantiate these trees by substituting individuals for \emph{semantic
  roles}, such as $\mathsf{self}$ and $\mathsf{subj}$, and then
combine the tree instances as shown in
Figure~\ref{fig:white-rabbit-sleeps-deriv} to obtain the sentence
``The white rabbit sleeps''.

@@

We can now consider various different algorithms for solving the SPUD
problem. The naive algorithm builds the TAG derivation top-down. In
the example, it starts with the elementary tree for ``sleeps''. This
tree satisfies the need to convey the semantic information, but
introduces a need to generate a noun phrase (NP) for the subject,
which must refer uniquely to the target referent $r_1$. In a second
step, we substitute the tree for ``the rabbit'' into the open NP leaf,
which makes the derivation grammatically complete. Since there are two
different individuals that could be described as ``the
rabbit''---technically, $r_2$ is still a \emph{distractor} (i.e.,
based on the description ``the rabbit'', the hearer might erroneously
think that we're talking about $r_2$ and not $r_1$)---we are still not
finished. To complete the derivation, the tree for ``white'' is added
to the existing structure by an \emph{adjunction} operation, making
the derivation both syntactically and semantically complete. This
search algorithm takes worst-case exponential time because it may have
to consider many combinations of elementary trees; and in fact, the
SPUD problem generalizes the surface realization problem considered by
\citet{KolStr02} and is therefore NP-complete.

@@

The SPUD system itself worked around this complexity by using a
greedy, incomplete search algorithm. On the other hand,
\citet{KolSto07} present an alternative to controlling the necessary
search, by encoding the generation problem into a planning
problem.\footnote{See \url{http://code.google.com/p/crisp-nlg/} for
  the CRISP system, in which this conversion is implemented.} Each
operator of this planning problem models the addition of some
elementary tree to the TAG derivation. For instance,
Figure~\ref{fig:white-rabbit-as-planning} shows the corresponding PDDL
actions for the above generation task. The \texttt{add-sleeps}
operator represents the addition of the elementary tree for ``sleeps''
to the derivation. $\verb?add-sleep?(\mathsf{root},e,r_1)$ removes the
atom $\mathsf{subst}(S,\mathsf{root})$, which expresses that there is
currently an open substitution node $\mathsf{root}$ with label $S$,
from the planning state and replaces it with an atom
$\mathsf{subst}(NP,\mathsf{subj}(\mathsf{root}))$; here
$\mathsf{subj}(\mathsf{root})$ is meant as shorthand for a fresh
individual name.\footnote{These terms, which are not valid in ordinary
  PDDL, can be eliminated by estimating an upper bound $n$ for the
  plan length, making $n$ copies of each action, ensuring that copy
  $i$ can only be applied in step $i$, and replacing the term
  $\mathsf{subj}(u)$ in an action copy by the constant
  $\mathsf{subj}_i$. Notice that $S$, $NP$, and $N$ are constants.}
At the same time, the operator records that the semantic information
$\mathsf{sleep}(e,r_1)$ has now been expressed, and introduces all
individuals except for $r_1$ as distractors for the new RE at
$\mathsf{subj}(\mathsf{root})$. These distractors can then be removed
by subsequent applications of the other two operators. By assuming an
initial state that encodes the knowledge base and contains an atom
$\mathsf{subst}(S,\emph{root})$, and a goal that contains, among
others, $\forall x \forall y. \neg \mathsf{subst}(x,y)$ and
$\mathsf{expressed}(\emph{sleep},e,r_1)$, a planner will find plans
like the following:
%
\begin{enumerate}
\item $\mathsf{add}\textsf{-}\mathsf{sleeps}(\mathsf{root}, r_1)$,
\item $\mathsf{add}\textsf{-}\mathsf{rabbit}(\mathsf{subj}(\mathsf{root}),r_1)$,
\item $\mathsf{add}\textsf{-}\mathsf{white}(\mathsf{subj}(\mathsf{root}),r_1)$.
\end{enumerate}
%
Using this plan, the grammatical derivation in
Figure~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the
generated sentence ``the white rabbit sleeps'', can be systematically
reconstructed. Thus, we can solve the sentence generation problem via
the detour through planning and bring current search heuristics for
planning to bear on generation.

>>>>

One way of modelling the sentence generation problem is to assume a
lexicalized grammar in which each lexicon entry specifies how it can
be combined grammatically with the other lexicon entries, what piece
of meaning it expresses, and what the pragmatic conditions on using it
are. Sentence generation can then be seen as constructing a
grammatical derivation that is syntactically complete, respects the
semantic and pragmatic conditions, and achieves all the
\emph{communicative goals}. An example is the tree-adjoining grammar
(TAG; \citep{joshi;etal1997}) shown in
Figure~\ref{fig:white-rabbit-sleeps-grammar}. This grammar consists of
\emph{elementary trees} (i.e., the disjoint trees in the figure), each
of which contributes certain \emph{semantic content}. Let's say that
we assume a knowledge base containing the individuals $e$, $r_1$ and
$r_2$ and the facts that $r_1$ and $r_2$ are rabbits, $r_1$ is white
and $r_2$ is brown, and $e$ is an event in which $r_1$ sleeps. Then we
would want to construct a sentence expressing the information
$\{\mathsf{sleep}(e,r_1)\}$ by combining instances of the elementary
trees (in which the \emph{semantic roles}, such as $\mathsf{self}$ and
$\mathsf{subj}$, have been substituted by constants from the knowledge
base) into a TAG derivation as as shown in
Figure~\ref{fig:white-rabbit-sleeps-deriv}. We can then read off the
sentence ``The white rabbit sleeps'' from the derivation. \todo{need
  to mention distractors and unique REs here}

@@

This perspective on sentence generation has the advantage of solving
the sentence planning and surface realization problems simultaneously:
in the example, we can require that referring expressions ``the white
rabbit'' can be resolved uniquely to $r_1$ by the hearer, in addition
to the requirement that the derivation is grammatically correct. This
is useful in cases where sentence planning and surface realization
interact, because syntactic information about the individual words is
available when the REs are generated (see
e.g. \cite{stone98textual}). However, the problem of whether a given
communicative goal can be achieved with a given grammar is NP-complete
\cite{KolStr02}. A naive search algorithm that computes the derivation
top-down takes exponential time and is clearly infeasible to use in
practice. The SPUD system \cite{Stone2003a}, which worked this idea
out for the first time, uses a greedy algorithm, which circumvents the
combinatorial explosion but is incomplete.

@@

Here we will look at one recent approach which aims at controlling the
search by converting the sentence generation problem into a planning
problem and then running a planner \citet{KolSto07}.\footnote{See
  \url{http://code.google.com/p/crisp-nlg/} for the CRISP system, in
  which this conversion is implemented.} \todo{XXX} Each operator of this
planning problem models the addition of some elementary tree to the
TAG derivation. For instance,
Figure~\ref{fig:white-rabbit-as-planning} shows the corresponding PDDL
actions for the above generation task. The \texttt{add-sleeps}
operator represents the addition of the elementary tree for ``sleeps''
to the derivation. $\verb?add-sleep?(\mathsf{root},e,r_1)$ removes the
atom $\mathsf{subst}(S,\mathsf{root})$, which expresses that there is
currently an open substitution node $\mathsf{root}$ with label $S$,
from the planning state and replaces it with an atom
$\mathsf{subst}(NP,\mathsf{subj}(\mathsf{root}))$; here
$\mathsf{subj}(\mathsf{root})$ is meant as shorthand for a fresh
individual name.\footnote{These terms, which are not valid in ordinary
  PDDL, can be eliminated by estimating an upper bound $n$ for the
  plan length, making $n$ copies of each action, ensuring that copy
  $i$ can only be applied in step $i$, and replacing the term
  $\mathsf{subj}(u)$ in an action copy by the constant
  $\mathsf{subj}_i$. Notice that $S$, $NP$, and $N$ are constants.}
At the same time, the operator records that the semantic information
$\mathsf{sleep}(e,r_1)$ has now been expressed, and introduces all
individuals except for $r_1$ as distractors for the new RE at
$\mathsf{subj}(\mathsf{root})$. These distractors can then be removed
by subsequent applications of the other two operators. By assuming an
initial state that encodes the knowledge base and contains an atom
$\mathsf{subst}(S,\emph{root})$, and a goal that contains, among
others, $\forall x \forall y. \neg \mathsf{subst}(x,y)$ and
$\mathsf{expressed}(\emph{sleep},e,r_1)$, a planner will find plans
like the following:
%
\begin{enumerate}
\item $\mathsf{add}\textsf{-}\mathsf{sleeps}(\mathsf{root}, r_1)$,
\item $\mathsf{add}\textsf{-}\mathsf{rabbit}(\mathsf{subj}(\mathsf{root}),r_1)$,
\item $\mathsf{add}\textsf{-}\mathsf{white}(\mathsf{subj}(\mathsf{root}),r_1)$.
\end{enumerate}
%
Using this plan, the grammatical derivation in
Figure~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the
generated sentence ``the white rabbit sleeps'', can be systematically
reconstructed. Thus, we can solve the sentence generation problem via
the detour through planning and bring current search heuristics for
planning to bear on generation.
